---
title: "Assignment 1 - Language Development in ASD - part 2"
author: "Simon Hansen"
date: "14/09-2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Language development in Autism Spectrum Disorder (ASD)

Background: Autism Spectrum Disorder is often related to language impairment. However, this phenomenon has not been empirically traced in detail: i) relying on actual naturalistic language production, ii) over extended periods of time. We therefore videotaped circa 30 kids with ASD and circa 30 comparison kids (matched by linguistic performance at visit 1) for ca. 30 minutes of naturalistic interactions with a parent. We repeated the data collection 6 times per kid, with 4 months between each visit. We transcribed the data and counted: 
i) the amount of words that each kid uses in each video. Same for the parent.
ii) the amount of unique words that each kid uses in each video. Same for the parent.
iii) the amount of morphemes per utterance (Mean Length of Utterance) displayed by each child in each video. Same for the parent. 

This data is in the file you prepared in the previous class. 

NB. A few children have been excluded from your datasets. We will be using them next week to evaluate how good your models are in assessing the linguistic development in new participants.

We then want to test the language trajectory of child and parent over time.

This Markdown is structured in the following way:

1. The exercises: read them carefully. Under each exercise you will have to write your answers, once you have written and run the code. This is the part that you have to directly send to the teachers.
2. An (optional) guided template full of hints for writing the code to solve the exercises. Fill in the code and the paragraphs as required. Then report your results under the exercise part.
3. In exercise 4 you will be asked to create the best possible model of language development in TD and ASD children, picking and choosing whatever additional variables you want from the dataset. Next time, the models produced by the different groups will compete against each other to see who can produce the best model, so choose carefully!

You will have to have a github repository for the code and send the answers to Celine and Riccardo without code (but a link to your github repository). This way we can check your code, but you are also forced to figure out how to report your analyses :-)

N.B. The following lines are a summary of the questions to be answered, the step-by-step instructions and tips are below.

## Exercise 1) Preliminary Data Exploration

Describe the participant samples in the dataset (e.g. by diagnosis, age, etc.). Do you think the two groups are well balanced? If not, what do you think was the reason?

[HERE GOES YOUR ANSWER]

According to the test conducted the only concern seem to be the age difference. So in general the two groups seem to be well balanced. 

### Exercise 2) Children learning language: the effects of time and ASD
Describe linguistic development in TD and ASD children in terms of Mean Length of Utterance (MLU)?

[HERE GOES YOUR ANSWER]

The model suggests a positive relationship between the number of visits and the MLU of the child, b= 0.23, t(60.78) = 9.52, p < .05

According to the model there is no significant influence of the diagnosis on the MLU of the child, b=0.29, (60.08) = 1.91, p > 0.5

So it seems like time has an influence on the child MLU whereas the diagnosis of the child does not influence the child MLU significantly.

### Exercise 3) Child directed speech as a moving target
Describe how parental use of language changes over time in terms of MLU. What do you think is going on?

[HERE GOES YOUR ANSWER]

According to the model there is a positive relationship of the number of visits on the MLU of the parents, b = 0.12, t(59.53) = 6.60, p < .05. This seems very reasonable that the mother will start using more complex sentences as the child grows older. 

The model also suggest that the parents of the typically developing children have significantly larger MLU's than those who are parent of autistic children, b= 0.50, t(60.92) = 4.43, p < .05. A possible explanation for this phenomenoum could be that parents of autistic children think that their children have a harder time acquiring language and therefor they try to help them by speaking in shorter utterances that are easier to comprehend. 

### Exercise 4) Looking into "individual differences" (demographic, clinical or cognitive profiles)
The dataset contains some additional variables characterizing the kids’ cognitive and clinical profile: ADOS (autism severity), MSEL EL (Expressive Language, that is, verbal IQ, or linguistic skills at first visit as assessed by a psychologist using Mullen Scales of Early Learning), MSEL VR (Visual Reception, used as a proxy for non verbal IQ at first visit), Age, Gender, Ethnicity. Would it make sense to add any of them to your model of linguistic trajectories? Create the best possible model (the one that best explain the data, with MLU as outcome). Next time your model will be tested on new participants, and we will proclaim a winner. Describe your strategy to select the best models (how did you choose the variables to include?) and send the code to Riccardo and Celine.

[HERE GOES YOUR ANSWER]

In our model we have chosen to predict the MLU of the children using the variables: Visit, mlu of the mother and the verbal IQ. The decision of using excatly these variables were motivated by theory and by testing different models to see which one would have the most explanatory power.

### [OPTIONAL] Exercise 5) Comment on how the three linguistic variables measure linguistic performance (the so-called "construct validity" of the measures). Do they express the same variance?

[HERE GOES YOUR ANSWER]

MLU: MLU is the mean length of utterance and therefore is an indicator of the childs ability to construct sentences. The longer the mlu the more advanced a setence can be made.

Types: Types is a measure of how many unique words the child uses during one session. This tells us something about the size of the vocabulary that the child posesses. The more words the child knows the easier it is to engange in different kinds of conversation and it might also be easier to specify things (such as using "owl" instead of just "bird")

Tokens: Tokens is a meaure of how many words the child uses in total during one session. This variable might be the least informative one in regards to linguistic skills as the child could say "cat" 200 times out of a total of 250 words. However this variable can tell us something about how much the child speaks in general.

While these three linguistic variables might often be correlated they tell us about different aspects of linguistic skills. A child might well know a lot of words, but be unable to put them together into long utterances. Therefor using all three variables can give us a nuanced way of asessing a child's level of linguistic skill.  

### Structure of the code chunks

Basic stuff:
- Loading the libraries
- Setting the directory and loading the data
- Look at the data (which variables are there? Are they in the right format?) and describe the participants (by diagnosis)

We will try to answer three questions:

- Do children with ASD develop language differently from non-ASD children?
- Do parents speak differently to children with ASD than to non-ASD ones?
- Which variables should we use to best explain the child linguistic performance?
  
### Loading the relevant libraries

Load necessary libraries : what will you need?

- e.g. something to plot with
- e.g. mixed effects models

```{r Load Libraries}
library(ggplot2)
library(lmerTest)
library(MuMIn)
```

### Define your working directory and load the data

- Create a new variable called locpath (localpath)
- Set it to be equal to your working directory
- Move to that directory (setwd(locpath))
- Load the data you saved last time (use read_csv(fileName))

```{r Load Data}

setwd("C:/Users/simon/Google Drev/Uni/Methods3/Assignment_1")
Data = read.csv("clean_data.csv")

# To omit first column
Data = Data[,-1]
```

### Characterize the participants (Exercise 1)

Identify relevant variables: participants demographic characteristics, diagnosis, ADOS, Verbal IQ, Non Verbal IQ, Visit, Number of words used, Number of unique words used, length of utterance in both child and parents.

Make sure the variables are in the right format.

Describe the characteristics of the two groups of participants and whether the two groups are well matched.

```{r}
# To get data only for first visit
data_visit1=subset(Data, Data$VISIT == 1)

# To make a T-test of age
m1=lm(Age ~ Diagnosis, data_visit1)
summary(m1)

ggplot(data_visit1, aes(x= Diagnosis, y= Age)) + geom_violin()

# To make Chi-square test of Gender
m2=glm(Gender ~ Diagnosis, family = binomial, data_visit1)
summary(m2)

# To make T-test for verbal IQ
m3=lm(verbalIQ ~ Diagnosis, data_visit1)
summary(m3)

ggplot(data_visit1, aes(x= Diagnosis, y= verbalIQ)) + geom_violin()

# To make T-test for non-verbal IQ
m4=lm(nonVerbalIQ ~ Diagnosis, data_visit1)
summary(m4)

ggplot(data_visit1, aes(x= Diagnosis, y= nonVerbalIQ)) + geom_violin()

```

[REPORT THE RESULTS]

Age: ASD children are significantly older than TDs, b = -12.63, t(60)=-12.32, p < .05

Gender: Our results suggest that gender has a non-signficant influence on the diagnosis, b= -0.37 z(60) = -0.52 p > .05

non-verbal IQ: Our results suggest that there is a non-significant difference between the non-verbal IQ of TDs and ASDs, b = -0,90 t(60) = -0.76, p > .05

Verbal IQ: Our results suggest that there is a non-significant difference between the verbal IQ of TDs and ASDs, b = 2.91, t(60) = 1.79, p > .05


## Let's test hypothesis 1: Children with ASD display a language impairment  (Exercise 2)

### Hypothesis: The child's MLU changes: i) over time, ii) according to diagnosis

Let's start with a simple mixed effects linear model

Remember to plot the data first and then to run a statistical test.
- Which variable(s) should be included as fixed factors?
- Which variable(s) should be included as random factors?

```{r}
# To create a plot which makes us able to get an overview of the data. NOTE: The plot does not represent the model!
ggplot(Data, aes(y=CHI_MLU, x= VISIT, colour = Diagnosis)) + geom_jitter() + geom_smooth(method = lm) 

# To make a mixed effects model
m5 = lmer(CHI_MLU ~ 1 + VISIT + Diagnosis + (1 + VISIT|ID), Data, REML = FALSE)
summary(m5)

```

How would you evaluate whether the model is a good model?

```{r}
# By comparing the model to a null model using anova and getting the r squared for the model

# To make null model
m_null = lmer(CHI_MLU ~ 1 + VISIT + (1 + VISIT|ID), Data, REML = FALSE)

# To test the model 
anova(m_null, m5)

# To get R squared
r.squaredGLMM(m5)

```

Not too good, right? Let's check whether a growth curve model is better.
Remember: a growth curve model assesses whether changes in time can be described by linear, or quadratic, or cubic (or... etc.) components.
First build the different models, then compare them to see which one is better

```{r}



```

Exciting right?
Now it's time to report our results.
Remember to report:
- the estimates for each predictor (beta estimate, standard error, p-value)
- A plain word description of the results

[REPORT THE RESULTS]
Linguistic development of children MLU is affected by ... [COMPLETE]

## Let's test hypothesis 2: Parents speak equally to children with ASD and TD  (Exercise 3)

### Hypothesis: Parental MLU changes: i) over time, ii) according to diagnosis

```{r}

# To create a plot which makes us able to get an overview of the data. NOTE: The plot does not represent the model!
ggplot(Data, aes(y=MOT_MLU, x= VISIT, colour = Diagnosis)) + geom_jitter() + geom_smooth(method = lm) 

# To make a mixed effects model
m6 = lmer(MOT_MLU ~ 1 + VISIT + Diagnosis + (1 + VISIT|ID), Data, REML = FALSE)
summary(m6)

```

[REPORT THE RESULTS]

### Adding new variables (Exercise 4)

Your task now is to figure out how to best describe the children linguistic trajectory. The dataset contains a bunch of additional demographic, cognitive and clinical variables (e.g.verbal and non-verbal IQ). Try them out and identify the statistical models that best describes your data (that is, the children's MLU). Describe how you selected the best model and send the code to run the model to Riccardo and Celine.


```{r}
m7 = lmer(CHI_MLU ~ 1 + VISIT + MOT_MLU + verbalIQ + (1 + VISIT|ID), Data, REML = FALSE)
summary(m7)

r.squaredGLMM(m7)

```

REPORT OF THE RESULTS